---
title: "Assignment 5 - Airline Delays: Tidying and Transforming Data"
author: "Sarah Abdelrahman"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

This analysis examines arrival delays for Alaska Airlines and AM West
across five destinations: Los Angeles, Phoenix, San Diego, San Francisco,
and Seattle. The goal is to tidy and transform the original data and
compare the airlines' overall performance with their performance across
individual destinations.

The data are based on the airline-delay table from *Numbersense* by
Kaiser Fung, McGraw Hill, 2013.

## Load Packages and Data

The airline delay data were recreated from the source table in wide
format and stored in a publicly accessible GitHub repository.

The raw CSV file is available at:

<https://raw.githubusercontent.com/sarahcunyms2026-source/Week-5-Airline-Delays/refs/heads/main/airline_delays.csv>

```{r load-data}
library(tidyverse)

url <- "https://raw.githubusercontent.com/sarahcunyms2026-source/Week-5-Airline-Delays/refs/heads/main/airline_delays.csv"

airlines <- read.csv(
  url,
  na.strings = "",
  strip.white = TRUE
)

airlines
```

## Inspecting the Data

Before transforming the data, I checked its structure and examined it for
missing values. This helps verify that the data were imported correctly
and identifies any cleanup that may be required.

```{r inspect-data}
str(airlines)

colSums(is.na(airlines))
```

The inspection identified two missing values in the `Airline` column.
These correspond to the delayed rows in the original source table, where
the airline names were intentionally left blank. These blank cells were
preserved when recreating the CSV file.

## Handling Missing Data

To populate the missing airline names, I used `fill()` from the `tidyr`
package. This carries the airline name from the preceding row down into
the corresponding delayed row.

```{r missing-data}
airlines_clean <- airlines %>%
  fill(Airline, .direction = "down")

airlines_clean
```

I then checked the cleaned dataset to confirm that the missing values had
been populated successfully.

```{r check-clean-data}
colSums(is.na(airlines_clean))
```

After using `fill()`, no missing values remained in the dataset.

## Transforming the Data from Wide to Long Format

The original dataset is in wide format, with each destination stored in a
separate column. To create a tidy dataset for analysis, I used
`pivot_longer()` to combine the five destination columns into two
variables: `Destination` and `Flights`.

```{r wide-to-long}
airlines_long <- airlines_clean %>%
  pivot_longer(
    cols = Los_Angeles:Seattle,
    names_to = "Destination",
    values_to = "Flights"
  ) %>%
  mutate(
    Destination = str_replace_all(Destination, "_", " ")
  )

airlines_long
```

The resulting long-format dataset contains one row for each combination
of airline, flight status, and destination. This structure makes it easier
to group, summarize, and compare airline performance.

## Count Analysis

Before comparing percentages, I summarized the number of on-time and
delayed flights for each airline across all five destinations. This
provides an initial view of the overall flight volume and delay counts.

```{r count-analysis}
overall_counts <- airlines_long %>%
  group_by(Airline, Status) %>%
  summarise(
    Total_Flights = sum(Flights),
    .groups = "drop"
  )

knitr::kable(
  overall_counts,
  caption = "Overall Flight Counts by Airline and Status"
)
```

Alaska had 501 delayed flights and 3,274 on-time flights. AM West had
787 delayed flights and 6,438 on-time flights.

However, these counts should not be compared by themselves because
AM West operated substantially more flights overall than Alaska.
Therefore, percentages provide a more appropriate comparison.

## Overall Delay Percentage

Because the two airlines operated different numbers of flights, comparing
delay counts alone would not provide a fair comparison. Therefore, I
calculated the percentage of flights that were delayed for each airline.

```{r overall-percentage}
overall_delay <- airlines_long %>%
  group_by(Airline) %>%
  summarise(
    Total_Flights = sum(Flights),
    Delayed_Flights = sum(Flights[Status == "delayed"]),
    Delay_Percentage = Delayed_Flights / Total_Flights * 100,
    .groups = "drop"
  )

overall_delay_table <- overall_delay %>%
  mutate(
    Delay_Percentage = round(Delay_Percentage, 2)
  )

knitr::kable(
  overall_delay_table,
  caption = "Overall Delay Percentage by Airline"
)
```

### Overall Comparison

Although AM West had more delayed flights than Alaska (787 compared with
501), AM West also operated substantially more flights overall.

When the number of delayed flights is considered as a percentage of total
flights, Alaska had an overall delay rate of approximately 13.27%, while
AM West had an overall delay rate of approximately 10.89%.

Therefore, based only on the overall delay percentages, AM West appears
to have better overall arrival performance than Alaska.

### Overall Delay Percentage Visualization

```{r overall-delay-chart}
ggplot(
  overall_delay,
  aes(
    x = Airline,
    y = Delay_Percentage,
    fill = Airline
  )
) +
  geom_col(width = 0.6) +
  geom_text(
    aes(
      label = paste0(
        round(Delay_Percentage, 2),
        "%"
      )
    ),
    vjust = -0.5
  ) +
  labs(
    title = "Overall Delay Percentage by Airline",
    x = "Airline",
    y = "Delay Percentage"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )
```

## Delay Percentage by Destination

To determine whether the overall results are consistent across individual
destinations, I calculated the delay percentage for each airline within
each of the five destinations.

```{r city-percentages}
city_delay <- airlines_long %>%
  group_by(Airline, Destination) %>%
  summarise(
    Total_Flights = sum(Flights),
    Delayed_Flights = sum(Flights[Status == "delayed"]),
    Delay_Percentage = Delayed_Flights / Total_Flights * 100,
    .groups = "drop"
  )

city_delay
```

### City-by-City Comparison

To make the comparison easier to read, I transformed the results so that
the two airlines appear side by side for each destination.

```{r city-comparison-table}
city_comparison <- city_delay %>%
  select(
    Airline,
    Destination,
    Delay_Percentage
  ) %>%
  mutate(
    Delay_Percentage = round(Delay_Percentage, 2)
  ) %>%
  pivot_wider(
    names_from = Airline,
    values_from = Delay_Percentage
  )

knitr::kable(
  city_comparison,
  caption = "Delay Percentages by Destination"
)
```

The city-by-city comparison shows that Alaska had a lower delay
percentage than AM West in every destination.

Alaska's delay percentages were 11.09% in Los Angeles, 5.15% in Phoenix,
8.62% in San Diego, 16.86% in San Francisco, and 14.21% in Seattle.

In comparison, AM West's delay percentages were 14.43% in Los Angeles,
7.90% in Phoenix, 14.51% in San Diego, 28.73% in San Francisco, and
23.28% in Seattle.

Therefore, when the airlines are compared within each individual
destination, Alaska performed better in all five cities.

### Delay Percentage by Destination Visualization

```{r city-delay-chart}
ggplot(
  city_delay,
  aes(
    x = Destination,
    y = Delay_Percentage,
    fill = Airline
  )
) +
  geom_col(
    position = "dodge"
  ) +
  geom_text(
    aes(
      label = paste0(
        round(Delay_Percentage, 2),
        "%"
      )
    ),
    position = position_dodge(width = 0.9),
    vjust = -0.4,
    size = 3
  ) +
  labs(
    title = "Delay Percentage by Airline and Destination",
    x = "Destination",
    y = "Delay Percentage",
    fill = "Airline"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

## Overall vs. City-by-City Discrepancy

The results show an important discrepancy between the overall comparison
and the city-by-city comparison. Overall, AM West appears to perform
better, with a delay rate of approximately 10.89%, compared with 13.27%
for Alaska. However, when the airlines are compared separately within
each destination, Alaska has a lower delay percentage than AM West in
all five cities.

This apparent contradiction occurs because the two airlines do not have
the same distribution of flights across destinations. AM West operates
a very large proportion of its flights to Phoenix, where delay rates are
relatively low. In contrast, Alaska operates a large proportion of its
flights to Seattle, where delay rates are higher.

As a result, the overall percentages are influenced by the different
numbers of flights that each airline operates in each destination.

Therefore, the aggregated overall comparison makes AM West appear to
perform better even though Alaska performs better within every individual
destination. This reversal between the aggregated and subgroup results
is an example of **Simpson's Paradox**.

## Conclusion

This analysis demonstrates why percentages and subgroup comparisons are
important when evaluating airline performance.

Although AM West had the lower overall delay rate (10.89% compared with
13.27% for Alaska), Alaska had the lower delay rate in all five individual
destinations.

The difference occurs because the airlines have very different
distributions of flights across destinations. AM West operates a large
proportion of its flights to Phoenix, where delay rates are relatively
low, while Alaska operates a large proportion of its flights to Seattle,
where delay rates are higher. These differences in flight volume affect
the aggregated percentages.

The analysis also demonstrates the importance of tidying and transforming
data before drawing conclusions. The source data were recreated in wide
format while preserving the original missing cells. R was then used to
populate the missing airline names and transform the data from wide to
long format.

The tidy dataset made it easier to group the observations, calculate
counts and percentages, compare the airlines overall and by destination,
and identify the discrepancy between the aggregated and city-level
results.